---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
## Loading and preprocessing the data
```{r get data}
setwd("C:/Users/Dan/Dropbox/personal/Coursera/data science certificate/Reproducible research/course projects/RepData_PeerAssessment1")
## the directory where the assignment repository was cloned. 
## I was not able to read a csv file directly out of a zipped folder, so I make a new directory 
## instead 

## unzip("activity.zip", exdir="activity") ## no need to run again in knitr

activity <- read.csv("activity.csv", header=T)

## take a look at and process data

activityNames <- data.frame(names(activity))
description <- c("Number of steps taken in a 5-minute interval","Date of measurement","5-minute interval in which measurement was taken")
activityNames <- data.frame(cbind(activityNames,description))
activityNames

require(psych)
describe(activity)

require(dplyr)
tbl_df(activity)
```
## What is mean total number of steps taken per day?
For this part of the assignment, you can ignore the missing values in the dataset.

    Make a histogram of the total number of steps taken each day

```{r Total Steps per Interval}
byDay <- group_by(activity, date)
byDaySum <- summarise(byDay, TotalSteps=sum(steps,na.rm=T))

require(ggplot2)
ggplot(byDaySum, aes(TotalSteps)) + geom_histogram()
```

    Calculate and report the mean and median total number of steps taken per day

``` {r}
describe(byDaySum)
```
The mean is 9,354 and the median is 10,395

## What is the average daily activity pattern?
    Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
    
``` {r Average steps per interval}
byInterval <- group_by(activity,interval)
byIntervalMean <- summarise(byInterval, StepsPerInterval=mean(steps,na.rm=T))
ggplot(byIntervalMean, aes(interval, StepsPerInterval)) + geom_line()
```

    Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

``` {r}
summary(byIntervalMean)
describe(byIntervalMean)
arrange(byIntervalMean, desc(StepsPerInterval))
```
Interval #835
## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

    Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

``` {r}
count(activity[activity=="NA",])
```
2,304

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

``` {r}
activityImputed <- activity
activityImputed <- left_join(activityImputed, byIntervalMean)
for(i in 1:length(activityImputed$steps)) {
    if(is.na(activityImputed$steps[i]==T)) { (activityImputed$steps[i]=activityImputed$StepsPerInterval[i]) } }
```


Create a new dataset that is equal to the original dataset but with the missing data filled in.

This is the activityImputed data set established above. 

    Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

``` {r Total steps per interval - imputed}
byDayImputed <- group_by(activityImputed,date)
byDayImputedSum <- summarise(byDayImputed, TotalSteps=sum(steps))
ggplot(byDayImputedSum,aes(TotalSteps)) + geom_histogram()
summary(byDayImputedSum)
```
The mean and median are both 10,766. This is higher than the original dataset because there are more observations at the center (mean) of each interval's distribution.   
## Are there differences in activity patterns between weekdays and weekends?
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

    Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

``` {r}
activityImputed$date <- as.POSIXlt(activityImputed$date,format="%m/%d/%Y")
str(activityImputed$date)
activityImputed$day <- weekdays(activityImputed$date)
for(i in 1:length(activityImputed$day)) {
if(activityImputed$day[i]=="Saturday") { activityImputed$weekend[i]="Yes" }
if(activityImputed$day[i]=="Sunday") {activityImputed$weekend[i]="Yes" }
else activityImputed$weekend[i]="No" }
```
    Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:

``` {r Average steps per interval by weekday or weekend}
noDate <- select(activityImputed, -date)
byIntervalImputed <- group_by(noDate, interval)
weekend <- byIntervalImputed[byIntervalImputed$weekend=="Yes",]
weekday <- byIntervalImputed[byIntervalImputed$weekend=="No",]
weekendSteps <- summarise(weekend, StepsPerInterval=mean(steps))
weekdaySteps <- summarise(weekday, StepsPerInterval=mean(steps))
weekendPlot <- ggplot(weekendSteps, aes(StepsPerInterval)) + geom_histogram() + ggtitle("Weekend")
weekdayPlot <- ggplot(weekdaySteps, aes(StepsPerInterval)) + geom_histogram() + ggtitle("Weekday")
require(gridExtra)
grid.arrange(weekendPlot, weekdayPlot)
```

Definitely a difference. During the week you have movement that peaks in the morning then settles. During the weekend you have steady movement throughout the day. 
